In-silico molecular enrichment and clearance of the human intracranial space

### 2024-12-19 paper/supplementary_main.tex
Add a justification and discussion of the 1D no-flux condition.

In terms of geometrical domains, we consider the parenchyma
$\Omega_{\rm PAR} \subset \R^3$ and CSF spaces $\Omega_{\rm CSF}
\subset \R^3$ with $\Omega = \Omega_{\rm PAR} \cup \Omega_{\rm CSF}$
(see also \Cref{fig:intracranial_domains}). The coordinates in these
3D domains is denoted by $x$. The interface between the parenchyma and
CSF spaces is given by $\partial \Omega_{\rm PAR} \cap \partial
\Omega_{\rm CSF}$ and we separate it as two parts: the surface of the
lateral ventricles $\Gamma_{\rm LV}$ and the remaining pial interface
$\Gamma_{\rm pia}$ (thus $\partial \Omega_{\rm CSF} \bigcap \partial
\Omega_{\rm PAR} = \Gamma_{\rm LV} \cup \Gamma_{\rm pia}$). The
remaining, outer, boundary of $\Omega_{\rm CSF}$ is again separated
into two parts: $\Gamma_{\rm SSAS}$ represents the lower interface
towards the spinal subarachnoid space (SSAS), while $\Gamma_{\rm AM}$
is the outer interface towards the arachnoid and dura
membranes. $\Gamma_{\rm AM}$ is further subdivided into its lower and
upper parts: $\Gamma_{\rm AM-L}$ and $\Gamma_{\rm AM-U}$. The boundary
towards the spinal cord is denoted by $\Gamma_{\rm{SC}}$ and given by
$\Gamma_{\rm{SC}} = \partial \Omega_{\rm PAR} \setminus \partial
\Omega_{\rm CSF}$.

In addition, we consider two sets of perivascular networks: a
periarterial network $\Lambda_a$ represented by the (connected)
centerlines $\Lambda^i_{(a)}$ of the arterial tree, and a perivenous
network $\Lambda_v$ associated with the centerlines $\Lambda^i_{(v)}$
of the veins. Recall that we represent the vascular domains as the
union of cylindrical vessels of radius $R_1^i$ surrounding the
centerlines $\Lambda^i$. Moreover, we consider the surrounding
perivascular spaces as the union of annular cylinders $\Omega^i$ of
inner radius $R_1^i$ and outer radius $R_2^i > R_1^i$ and thus of
width $R_2^i - R_1^i$. We denote the outer lateral surface of the
periarterial and perivenous spaces by $\Gamma_a$ and $\Gamma_v$,
respectively. We omit the subscript $a$ or $v$ when referring to any
such network, vessel segment or perivascular outer surface. We assume
that $\Lambda$ is parametrized by the coordinate $s$, and with a minor
abuse of notation, simply write $s \in \Lambda$ to represent the point
${\bm \lambda}(s)$ on $\Lambda$ corresponding to $s$. Finally, we view
each perivascular network both as a geometric domain and as a directed
graph with the centerlines $\{ \Lambda^i \}$ as oriented edges and the
connections as nodes $\mathcal{V}$.

@@

In this section, we describe the mathematical
model~\eqref{eq:multi_transport}, the associated definitions,
interface and boundary conditions in further detail. We refer to the
main text for explicit parameter values while giving the general,
abstract form here, and to the reference~\cite{masri2023modelling} for
the derivation and analysis of this 3D-1D model. Recall that we model
a concentration field $c = c(x, t)$ for $x \in \Omega$ and $t > 0$
defined in $\Omega_{\rm PAR}$ and $\Omega_{\rm CSF}$ separately. In
each of these domains, $c$ satisfies
\mer{MER: Rami asks if it should be a $\phi$ in the advective term here.}
\begin{equation}
  \partial_t (\phi c) - \nabla \cdot (D \nabla (\phi c) ) + \nabla \cdot (\bm u c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma
  %= f
   = 0
  \quad \quad \mathrm{in} \quad \Omega_{\rm CSF}, \Omega_{\rm{PAR}}.
  \label{eq:3d_pde}
\end{equation}
In~\eqref{eq:3d_pde}, $\phi$ is the fluid volume fraction (also known as the porosity) defined in the parenchyma ($\phi \ll 1$ in $\Omega_{\rm PAR}$) and in the CSF spaces ($\phi = 1$ in $\Omega_{\rm CSF}$). In the parenchyma, $\phi$ represents the extracellular space volume fraction. Moreover, $D$ is the effective diffusion coefficient of the relevant solute in the respective media which takes different values over the CSF spaces and the parenchyma, depending on tortuosity~\cite{sykova2008diffusion} and dispersive effects; and $\bm u$ is a convective velocity field representing the flow of CSF in $\Omega_{\rm CSF}$ and the flow of ISF in $\Omega_{\rm PAR}$. $\xi$ models a transfer or exchange parameter between the 3D domain ($\Omega_{\rm PAR}$ or $\Omega_{\rm CSF}$) and the perivacular networks $\Lambda_a, \Lambda_v$. To summarize,
\begin{equation*}
\phi =  \begin{cases}
  1  & \mathrm{in} \;  \Omega_{\rm{CSF}} \\ 
  \phi_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
  \end{cases}, \; 
D = \begin{cases}
  D_{\rm{CSF}} & \mathrm{in} \;  \Omega_{\rm{CSF}} \\ 
  D_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} \end{cases}, \; 
  \quad \bm u  = \begin{cases}
  \bm{u}_{\rm{CSF}} & \mathrm{in} \; \Omega_{\rm{CSF}} \\ 
  \bm{u}_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
\end{cases}, \; 
\xi = \begin{cases}
  \xi_{\rm{CSF}} & \mathrm{if} |\Omega^i \cap \Omega_{\rm{CSF}}| \neq 0 \\
  \xi_{\rm{EF}} & \mathrm{if} |\Omega^i \cap \Omega_{\rm{PAR}}| \neq 0  
\end{cases} . 
\end{equation*}
In the above, $\xi$ is defined segment-wise: for each centerline $\Lambda^i$ with surrounding PVS $\Omega^i$, $|\Omega^i \cap \Omega_{\rm{PAR}}|$ (resp. $|\Omega^i \cap \Omega_{\rm{CSF}}|$) is nonzero whenever $\Omega^i$ intersects $\Omega_{\rm{PAR}}$ (resp. $\Omega_{\rm{CSF}}$). If the surrounding PVS intersects both, then we set $\xi = \xi_{\rm EF}$ if $\Omega^i$ mainly ($80$ percent) intersects $\Omega_{\rm{PAR}}$, in which case the interaction with $\Omega_{\rm{CSF}}$ is ignored; otherwise, we set $\xi = \xi_{\rm CSF}$.

@@

Finally, we assume no molecular efflux at all end nodes $\mathcal{N}$
via a homogeneous Neumann condition augmenting~\eqref{eq:1d_pde}:
\begin{equation}
    \hat D A \partial_s (\phi \hat c) - A \hat u \hat c  = 0 \quad \mathrm{on} \quad \mathcal{N}.
    \label{eq:1d_bcs}
\end{equation}

\mk{What is justification of \eqref{eq:1d_bcs}?}
%% Both $\Omega$ and $\Lambda$ are fixed in time and we assume
%% constant-in- time radii (despite varying radii to drive peristaltic
%% flow, see \cref{sec:sup:peristalsis}).

>>>>

In terms of geometrical domains, we consider the parenchyma
$\Omega_{\rm PAR} \subset \R^3$ and CSF spaces $\Omega_{\rm CSF}
\subset \R^3$ with $\Omega = \Omega_{\rm PAR} \cup \Omega_{\rm CSF}$
(\Cref{fig:results1}A, \Cref{fig:csf}A--C). The coordinates in these
3D domains is denoted by $x$. The interface between the parenchyma and
CSF spaces is given by $\partial \Omega_{\rm PAR} \cap \partial
\Omega_{\rm CSF}$ and we separate it as two parts: the surface of the
lateral ventricles $\Gamma_{\rm LV}$ and the remaining pial interface
$\Gamma_{\rm pia}$ (thus $\partial \Omega_{\rm CSF} \bigcap \partial
\Omega_{\rm PAR} = \Gamma_{\rm LV} \cup \Gamma_{\rm pia}$). The
remaining, outer, boundary of $\Omega_{\rm CSF}$ is again separated
into two parts: $\Gamma_{\rm SSAS}$ represents the lower interface
towards the spinal subarachnoid space (SSAS), while $\Gamma_{\rm AM}$
is the outer interface towards the arachnoid and dura
membranes. $\Gamma_{\rm AM}$ is further subdivided into its lower and
upper parts: $\Gamma_{\rm AM-L}$ and $\Gamma_{\rm AM-U}$. The boundary
towards the spinal cord is denoted by $\Gamma_{\rm{SC}}$ and given by
$\Gamma_{\rm{SC}} = \partial \Omega_{\rm PAR} \setminus \partial
\Omega_{\rm CSF}$.

In addition, we consider two sets of perivascular networks: a
periarterial network $\Lambda_a$ represented by the (connected)
centerlines $\Lambda^i_{(a)}$ of the arterial tree, and a perivenous
network $\Lambda_v$ associated with the centerlines $\Lambda^i_{(v)}$
of the veins. Recall that we represent the vascular domains as the
union of cylindrical vessels of radius $R_1^i$ surrounding the
centerlines $\Lambda^i$. Moreover, we consider the surrounding
perivascular spaces as the union of annular cylinders $\Omega^i$ of
inner radius $R_1^i$ and outer radius $R_2^i > R_1^i$ and thus of
width $R_2^i - R_1^i$. \discuss{We denote the outer lateral surface of the
periarterial and perivenous spaces by $\Gamma_a$ and $\Gamma_v$,
respectively.} We omit the subscript $a$ or $v$ when referring to any
such network, vessel segment or perivascular outer surface. We assume
that $\Lambda$ is parametrized by the coordinate $s$, and with a minor
abuse of notation, simply write $s \in \Lambda$ to represent the point
${\bm \lambda}(s)$ on $\Lambda$ corresponding to $s$. Finally, we view
each perivascular network both as a geometric domain and as a directed
graph with the centerlines $\{ \Lambda^i \}$ as oriented edges and the
connections as nodes $\mathcal{V}$.

@@

In this section, we describe the mathematical
model~\eqref{eq:multi_transport}, the associated definitions,
interface and boundary conditions in further detail. We refer to the
main text for explicit parameter values while giving the general,
abstract form here, and to the reference~\cite{masri2023modelling} for
the derivation and analysis of this 3D-1D model. Recall that we model
a concentration field $c = c(x, t)$ for $x \in \Omega$ and $t > 0$
defined in $\Omega_{\rm PAR}$ and $\Omega_{\rm CSF}$ separately. In
each of these domains, $c$ satisfies
%\mer{MER: Rami asks if it should be a $\phi$ in the advective term here. We can include it, or drop it, but in the latter case, perhaps we should clarify that u is an average/superficial/Darcy velocity.}
\begin{equation}
  \partial_t (\phi c) - \nabla \cdot (D \nabla (\phi c) ) + \nabla \cdot (\bm u c ) + \xi (\overline{c} - \hat c ) \delta_\Gamma
  %= f
   = 0
  \quad \quad \mathrm{in} \quad \Omega_{\rm CSF}, \Omega_{\rm{PAR}}.
  \label{eq:3d_pde}
\end{equation}
In~\eqref{eq:3d_pde}, $\phi$ is the fluid volume fraction (also known as the porosity) defined in the parenchyma ($\phi \ll 1$ in $\Omega_{\rm PAR}$) and in the CSF spaces ($\phi = 1$ in $\Omega_{\rm CSF}$). In the parenchyma, $\phi$ represents the extracellular space volume fraction, and thus $c$ here generally represents the intrinsic (in contrast to the superficial) concentration~\cite{sykova2008diffusion}. Moreover, $D$ is the effective diffusion coefficient of the relevant solute in the respective media which takes different values over the CSF spaces and the parenchyma, depending on tortuosity~\cite{sykova2008diffusion} and dispersive effects; and $\bm u$ is a convective velocity field representing the flow of CSF in $\Omega_{\rm CSF}$ and the flow of ISF in $\Omega_{\rm PAR}$. $\xi$ models a transfer or exchange parameter between the 3D domain ($\Omega_{\rm PAR}$ or $\Omega_{\rm CSF}$) and the perivacular networks $\Lambda_a, \Lambda_v$. To summarize,
\begin{equation*}
\phi =  \begin{cases}
  1  & \mathrm{in} \;  \Omega_{\rm{CSF}} \\ 
  \phi_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
  \end{cases}, \; 
D = \begin{cases}
  D_{\rm{CSF}} & \mathrm{in} \;  \Omega_{\rm{CSF}} \\ 
  D_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} \end{cases}, \; 
  \quad \bm u  = \begin{cases}
  \bm{u}_{\rm{CSF}} & \mathrm{in} \; \Omega_{\rm{CSF}} \\ 
  \bm{u}_{\rm{PAR}} & \mathrm{in} \; \Omega_{\rm{PAR}} 
\end{cases}, \; 
\xi = \begin{cases}
  \xi_{\rm{CSF}} & \mathrm{if} |\Omega^i \cap \Omega_{\rm{CSF}}| \neq 0 \\
  \xi_{\rm{EF}} & \mathrm{if} |\Omega^i \cap \Omega_{\rm{PAR}}| \neq 0  
\end{cases} . 
\end{equation*}
In the above, $\xi$ is defined segment-wise: for each centerline $\Lambda^i$ with surrounding PVS $\Omega^i$, $|\Omega^i \cap \Omega_{\rm{PAR}}|$ (resp. $|\Omega^i \cap \Omega_{\rm{CSF}}|$) is nonzero whenever $\Omega^i$ intersects $\Omega_{\rm{PAR}}$ (resp. $\Omega_{\rm{CSF}}$). If the surrounding PVS intersects both, then we set $\xi = \xi_{\rm EF}$ if $\Omega^i$ mainly ($80$ percent) intersects $\Omega_{\rm{PAR}}$, in which case the interaction with $\Omega_{\rm{CSF}}$ is ignored; otherwise, we set $\xi = \xi_{\rm CSF}$.

@@

Finally, we impose a homogogenous Neumann condition at all network end
nodes to augment~\eqref{eq:1d_pde}:
\begin{equation}
    \hat D A \partial_s (\phi \hat c) - A \hat u \hat c  = 0 \quad \mathrm{on} \quad \mathcal{N}.
    \label{eq:1d_bcs}
\end{equation}
This no-flux condition states that the network end nodes represent
barriers for molecular efflux into or out of the perivascular network,
and thus that all solute exchange between the PVSs and their
surroundings takes place via the lateral outer PVS surface and is
regulated by the exchange parameter $\xi$ cf.~\eqref{eq:3d_pde}
and~\eqref{eq:1d_pde}. Larger particles have been observed to
accumulate within the PVS as the surface arteries penetrate into the
brain parenchyma~\cite{mestre2018flow, bedussi2018paravascular},
and~\eqref{eq:1d_bcs} is appropriate to represent such
behavior. However, to represent a continuously extending PVSs also
along penetrating vessels, this condition should be revisited.

%\mk{What is justification of \eqref{eq:1d_bcs}?}
%% Both $\Omega$ and $\Lambda$ are fixed in time and we assume
%% constant-in- time radii (despite varying radii to drive peristaltic
%% flow, see \cref{sec:sup:peristalsis}).
